--- 
site: bookdown::bookdown_site
documentclass: book
bookdown::gitbook:
  css: style.css
---

# 2023 AARMS-EIDM Summer School {-}

```{r, include=F}
require(ggplot2)
require(cowplot)
require(magick)
require(patchwork)
g1=ggplot() + 
  draw_image("images/BBMS.jpg") 
g2=ggplot() +
  draw_image("images/lecture.jpg")
g3=ggplot() +
  draw_image("images/library.jpg")
```
```{r, echo=F}
g1+(g2/g3)
```


**The AARMS-EIDM Summer School will take place at [Bonne Bay Marine Station](https://grenfell.mun.ca/academics-and-research/Pages/Bonne-Bay-Marine-Station.aspx), located in Norris Point on the west coast of Newfoundland.** Two graduate-level mathematics courses will be offered from August 19 to 31, 2023.  [Photo credit](https://commons.wikimedia.org/wiki/File:Bonnebaymarinestation.jpg)

## Details {-}

Two courses will be offered:

- [Course 1.](#course1) Mathematical epidemiology, and

- [Course 2.](#course2) Data, models, and decision support.

Participants will take both courses and receive a grade and certificate of completion that may be used to transfer credit to the home institution. The intended audience for the courses are math graduate students who would like to take these courses for credit. The course is also suitable for undergraduates with an interest in mathematical biology (and who have completed the relevant prerequisite coursework), and advanced graduate students and postdoctoral fellows, who are interested in working collaboratively on an innovative project to advance infectious disease modelling.

Applicants should have completed undergraduate courses in mathematical modelling, dynamical systems, statistics, and be able to code in R and/or python.

From August 20 to 30, 7 hours of work are planned to occur at [Bonne Bay Marine Station](https://grenfell.mun.ca/academics-and-research/Pages/Bonne-Bay-Marine-Station.aspx). Participants should capitalize on being together in one location to work efficiently on projects. Due to the compressed schedule, participants will be asked to complete a problem set prior to arrival on August 19.

## Instructors {-}

- [Dr. Amy Hurford](https://ahurford.github.io/website/#dr.-amy-hurford), Memorial University (Organizer, email: ahurford@mun.ca)

- [Dr. Julien Arino](https://julien-arino.github.io/), University of Manitoba

- [Dr. Amy Greer](https://ovc.uoguelph.ca/population-medicine/faculty/amy-greer), University of Guelph

- [Dr. Jane Heffernan](https://news.yorku.ca/experts/?name_or_keyword=jane+heffernan&research_area=&submit=Search+the+Guide&search=Y), York University

- [Dr. James Watmough](https://watmough.ext.unb.ca/), University of New Brunswick

## Guest lecturers {-}

- [Dr. Bouchra Nasri](https://www.bouchrarnasri.com/), Universite de Montreal

- [Dr. Steve Walker](https://github.com/stevencarlislewalker), McMaster University

- [Dr. Brenda Wilson](https://www.mun.ca/medicine/faculty-and-staff-resources/faculty-a-z/wilson-brenda.php), Memorial University

- [Dr. Edward Thommes](https://sites.google.com/view/ewthommes/home), Sanofi Pasteur

## Funding {-}
```{r, include=F}
g1=ggplot() + 
  draw_image("images/AARMS.png") 
g2=ggplot() +
  draw_image("images/mfph.png")
g3=ggplot() +
  draw_image("images/canmod.png")
g4=ggplot() +
  draw_image("images/omni.png")
g5=ggplot() +
  draw_image("images/mun2.jpg")
```

Summer school participants will have their accommodation and meal costs covered. Accommodations will be booked for participants. Participants are responsible for organizing their own transportation to Bonne Bay Marine Station.

The summer school is supported by the [Atlantic Association for Research in the Mathematical Sciences](https://aarms.math.ca/), [Mathematics for Public Health](http://www.fields.utoronto.ca/activities/public-health), the [Canadian Network for Modelling Infectious Diseases](https://canmod.net/), the [One Health Modelling Network for Emerging Infections](https://omni-reunis.ca/), [Memorial University](https://grenfell.mun.ca/academics-and-research/Pages/Bonne-Bay-Marine-Station.aspx), and the [Canadian Centre for Disease Modelling](https://www.yorku.ca/science/cdm/).

```{r, echo=F, fig.width=6, fig.height=1}
layout <- '
ABCDE#
'
wrap_plots(A = g1, B = g2, C = g3,D=g4,E=g5, design = layout)
```